---
Title: Reproducable Research - Assign. 2
author: "Abhishek Kaul"
date: "Sunday, March 22, 2015"
output:
  html_document:
    keep_md: yes
---

#Synopsis

Based on the data analysis
- across the United States, tornado are most harmful with respect to population health. 
- across the United States floods have the greatest economic consequences


#Data Processing 

This section provides the data processing activities done to the data set. 

```{r, cache=TRUE}
library(ggplot2)

# read the dat file for storm data
data <- read.csv("repdata-data-StormData.csv")

# get familiar with the data 
names(data)
head(data,n=2)
summary(data)
table(complete.cases(data))

# initial plots to understand the data
ggplot(yearagg,aes(damage,event)) + geom_point() + geom_line()

# get colum names to lowercase
names(data) <- tolower(names(data))
```

## Now lets find the population harm for events
```{r}
# harmful for population is fatalaties and injuries ... aggregate & order them
fatal <- aggregate(fatalities ~ evtype, data = data, FUN = sum)
fatal <- fatal[order(fatal$fatalities, decreasing = T), ]
ggplot(fatal[1:7, ], aes(evtype, fatalities)) + geom_bar(stat = "identity") + ggtitle("Top Seven Types of Events Causing Deaths Across the U.S")

injury <- aggregate(injuries ~ evtype, data = data, FUN = sum)
injury <- injury[order(injury$injuries, decreasing = T), ]
ggplot(fatal[1:7, ], aes(evtype, fatalities)) + geom_bar(stat = "identity") + ggtitle("Top Seven Types of Events Causing Injuries Across the U.S")

```

This shows that top event for population harm is Tornado

## Now lets move to the economic damage 
```{r}
# check units
table(data$propdmgexp)
table(data$cropdmgexp)

# normalise 

pd <- data$propdmg
pde <- data$propdmgexp
cd <- data$cropdmg
cde <- data$cropdmgexp

pd[pde %in% "B"] <- pd[pde %in% "B"] * 1000
pd[pde %in% c("M", "m")] <- pd[pde %in% c("M", "m")] * 1
pd[pde %in% c("K")] <- pd[pde %in% c("K")] * 0.001
pd[pde %in% c("H", "h")] <- pd[pde %in% c("H", "h")] * 1e-04
pd[!(pde %in% c("B", "M", "m", "K", "H", "h"))] <- pd[!(pde %in% c("B", "M", 
    "m", "K", "H", "h"))] * 1e-06

cd[cde %in% "B"] <- cd[cde %in% "B"] * 1000
cd[cde %in% c("M", "m")] <- cd[cde %in% c("M", "m")] * 1
cd[cde %in% c("K", "k")] <- cd[cde %in% c("K", "k")] * 0.001
cd[!(cde %in% c("B", "M", "m", "K", "k"))] <- cd[!(cde %in% c("B", "M", "m", 
    "K", "k"))] * 1e-06

totaldamage <- cd + pd
edt <- aggregate(totaldamage ~ data$evtype, FUN = sum)
oedt <- edt[order(edt$totaldamage, decreasing = T), ]
names(oedt)[1] <- "evtype"

ggplot(oedt[1:7, ], aes(evtype, totaldamage)) + geom_bar(stat = "identity") + ggtitle("Top Seven Types of Events Causing Economic Damages Across the U.S")

```


#Results 


Based on the data analysis
- across the United States, tornado are most harmful with respect to population health (both injury & fatal. 
- across the United States floods have the greatest economic consequences for property & agriculture


